Fail Faster

### 2025-03-20 paper/paper.tex
done through 3.4

Returning to our running example, Figure~\ref{fig:running-staged} shows the
int-pair example written with the staged \texttt{Gen.t} monad, as well as the
inlined code that results from calling \texttt{Gen.to\_bq} (changing some identifier names for clarity).
The code generated is identical to the manually inlined version from Section~\ref{section:motiv}, and of course runs equally fast.

@@

\jwc{Is there anything more we need to say here?}

@@

\begin{figure}
\begin{lstlisting}
module Gen =
$\dots$
  let pick (acc : int code) (weighted_gens : (int code * 'a t) list) (size : int code) (random : Sr.t code) : 'a code =
    match weighted_gens with
    | [] -> .< failwith "Error" >.
    | (wc,g) :: gens' ->
      .<
        if .~acc <= .~wc then .~(g size random)
        else
          let acc' = .~acc - .~wc in
          .~(pick .<acc'>. gens' size random)
      >.

@@

Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
generators and weights. The weights themselves might only be known at run time --- it is
common to use the current size parameter as a weight, for instance ---
so they are \texttt{code}s.
\texttt{Gen.weighted\_union} begins by computing \texttt{sum\_code}, an \texttt{int code}
that is the sum of the weights. Note that this happens at compile time: we fold over a list
known at compile time to produce another code value. 
We then call \texttt{SR.int} to sample a random number \texttt{r} between \texttt{0} and the sum.
Finally, we splice in the result of calling the helper function \texttt{pick}. \texttt{pick}
produces a tree of \texttt{if}s by again traversing the list of generators at compile time.
This tree of \texttt{ifs} ``searches'' for the generator corresponding to the
sampled value \texttt{r}, and then runs it.
\hg{I'm not sure this helps me. I think I'd prefer a less detailed explanation
of the code that conveys the intuition and let the reader actually read the code
if they want to know specifics. As it stands the explanation is just too dense
for me to process}

Figure~\ref{fig:staged-weighted-union-example} demonstrates a use of this staged weighted union.
Given a list (in this case constant) generators with weights ``the current size parameter'', \texttt{2}, and \texttt{1},
the generated code first computes the sum of these numbers, samples between \texttt{0} and the sum, and then
traverses a tree of three \texttt{if}s to find the correct value to return.

>>>>

Returning to our running example, Figure~\ref{fig:running-staged} shows the
int-pair generator written with the staged \texttt{Gen.t} monad, as well as the
inlined code that results from calling \texttt{Gen.to\_bq} (changing some identifier names for clarity).
The code generated is identical to the manually inlined version from Section~\ref{section:motiv}, and of course runs equally fast.

@@

@@

\begin{figure}
\begin{lstlisting}
module Gen =
$\dots$
  let pick (acc : int code) (weighted_gens : (int code * 'a t) list) size random : 'a code =
    match weighted_gens with
    | [] -> .< failwith "Error" >.
    | (wc,g) :: gens' ->
      .<
        if .~acc <= .~wc then .~(g size random)
        else
          let acc' = .~acc - .~wc in
          .~(pick .<acc'>. gens' size random)
      >.

@@

Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
generators and weights. The weights themselves might only be known at run time --- it is
common to use the current size parameter as a weight, for instance ---
so they are \texttt{code}s.
Instead of building a histogram of the distribution described by the weights at run time and then traversing it,
the compile time \texttt{weighted\_union} combinator generates a
tree of \texttt{if}s --- specialized to the list of weights known at compile time --- that searches for the selected generator.

% \texttt{Gen.weighted\_union} begins by computing \texttt{sum\_code}, an \texttt{int code}
% that is the sum of the weights. Note that this happens at compile time: we fold over a list
% known at compile time to produce another code value. 
% We then call \texttt{SR.int} to sample a random number \texttt{r} between \texttt{0} and the sum.
% Finally, we splice in the result of calling the helper function \texttt{pick}. \texttt{pick}
% produces a tree of \texttt{if}s by again traversing the list of generators at compile time.
% This tree of \texttt{ifs} ``searches'' for the generator corresponding to the
% sampled value \texttt{r}, and then runs it.
% \hg{I'm not sure this helps me. I think I'd prefer a less detailed explanation
% of the code that conveys the intuition and let the reader actually read the code
% if they want to know specifics. As it stands the explanation is just too dense
% for me to process}

Figure~\ref{fig:staged-weighted-union-example} demonstrates a use of this staged weighted union.
Given a list (in this case constant) generators with weights ``the current size parameter'', \texttt{2}, and \texttt{1},
the generated code first computes the sum of these numbers, samples between \texttt{0} and the sum, and then
traverses the tree of three \texttt{if}s to find the correct value to return.
